P3064R2: How to Avoid OOTA Without Really Trying

### 2024-08-19 paper/ootavacuous.tex
paper: Remove Appendix E and references to it.
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

All OOTA workers owe a debt to the foundational work in the infamous
``Causality Test Cases'',\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
a version of which may be found in
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.

@@

Later examples will usually combine the litmus test and the \co{herd7}
output into one listing, as shown in
Listing~\ref{lst:Simple Reordering as Litmus Test},
which recasts
Listing~\ref{lst:Simple Reordering}
into litmus-test form.
Listing~\ref{lst:Simple OOTA}
can also be recast as a \co{herd7} litmus test, as shown in
Listing~\ref{lst:Causality Test Case 4}
on page~\pageref{lst:Causality Test Case 4}.
Other OOTA-related litmus tests may be found in
Appendix~\ref{app:Illustrative Litmus Tests}
and
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.

@@

This paper focuses primarily on compilers that do only local per-thread
analysis.
Further explorations could consider additional classes of global analysis,
for example, analyses that demonstrate that a given expression will
always evaluate to a particular constant.
JMM Causality Test Case 1 discussed in
Appendix~\ref{app:Causality Test Case 1}
provides an example of such a global analysis.

@@

The first step in this direction is
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
which expands
Listing~\ref{lst:Causality Test Case 4}
(discussed in
Appendix~\ref{app:Causality Test Case 4})
from two threads to three.
The \co{herd7} tool reports OOTA values, and
all threads have straightforward semantic dependencies from their
loads to their stores.

@@

\section{Litmus Tests from “Causality Test Cases"}
\label{app:Litmus Tests from “Causality Test Cases"}

These tests might be two decades old, but they are still quite
relevant.\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
We have translated them from Java to C++, and we evaluate them using
the \co{herd7} tool and also using manual analysis.
Please note that differences in the semantics of the two languages
result in changes in verdict for some litmus tests.

\subsection{Causality Test Case 1}
\label{app:Causality Test Case 1}

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
And indeed, this result does show up in the output from the \co{herd7} tool.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-1.litmus @@
\caption{Causality Test Case 1}
\label{lst:Causality Test Case 1}
\end{listing}

Compilers that can prove the value of \co{x} is always
nonnegative can also determine that there is no semantic dependency between
\co{P0()}'s load from \co{x} on line~8 and its store to \co{y} on line~10,
that is, no matter which of the possible values is loaded from \co{x},
the value 1 will always be stored to \co{y}.
Relative to less-omniscient compilers, however,
including those using single-thread analysis or treating atomic
objects as volatile, there \emph{is} a semantic dependency.

Even omniscient compilers will observe a semantic dependency from
\co{P1()}'s load from \co{y} on line~14 to its store to \co{x} on line~15.

Either way no OOTA cycle will occur, with omniscient compilers
having no cycle at all because of the lack of an sdep link in \co{P0()},
and other compilers prevented by hardware
dependency ordering from realizing the cycle.

\subsection{Causality Test Case 2}
\label{app:Causality Test Case 2}

Listing~\ref{lst:Causality Test Case 2}
shows causality test case 2, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
This result appears in the \co{herd7} output.

Because \co{P0()}'s loads from \co{x} on lines~8 and~9 are unordered,
a C++ compiler that treats atomics as quasi volatile
can act as if the source code loaded from \co{x} only
once and assigned the returned value to both \co{r1} and \co{r2},
merging the loads.\footnote{
	But not in C, where relaxed atomic operations are volatile,
	and thus are observable behavior not subject to merging.}
The compiler could then prove that the condition on line~10
would always true, and act as if the source code had
unconditionally executed the relaxed store to \co{y} on line~11
independent of the value loaded from \co{x}.
Without this dependency there would then be no OOTA cycle.

For compilers that do not merge the two loads, the hardware
dependency ordering arising from the conditional test on line~10
will prevent the realization of an OOTA cycle.

\subsection{Causality Test Case 3}
\label{app:Causality Test Case 3}

Listing~\ref{lst:Causality Test Case 3}
shows causality test case 3, for which the \co{r1 == r2 == r3 == 1} result
is to be allowed.
The \co{herd7} tool reports this result.

The analysis from the previous section applies here.
Although the addition of \co{P2()} allows sequentially consistent executions
in which \co{P0()}'s two loads from \co{x} return different values,
it does not change the fact that those two loads can be merged.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-3.litmus @@
\caption{Causality Test Case 3}
\label{lst:Causality Test Case 3}
\end{listing}

\subsection{Causality Test Case 4}
\label{app:Causality Test Case 4}

Listing~\ref{lst:Causality Test Case 4}
shows causality test case 4, for which the \co{r1 == r2 == 1} result
is to be forbidden.
The \co{herd7} tool finds this OOTA result;
the \co{S8} values given on line~3 of its output indicate
that in the execution it describes, the variables \co{r1} and~\co{r2} were never
assigned any specific value, so they could just as well be equal to
one as to anything else.

This test case is virtually the same as the Simple OOTA cycle
discussed in
Section~\ref{sec:Simple OOTA Cycle}
on
page~\pageref{sec:Simple OOTA Cycle},
the only difference being the final values of the variables.

\subsection{Causality Test Case 5}
\label{app:Causality Test Case 5}

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
result is to be forbidden.
The \co{herd7} tool reports this OOTA result in line~3 of its output.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-5.litmus @@
\caption{Causality Test Case 5}
\label{lst:Causality Test Case 5}
\end{listing}

This is the same as causality test case~4 with \co{P2()} and \co{P3()}
added, but those routines play no role in the OOTA cycle.
\co{P2()} has no loads and so cannot participate in an OOTA cycle in
any case.

\subsection{Causality Test Case 6}
\label{app:Causality Test Case 6}

Listing~\ref{lst:Causality Test Case 6}
shows causality test case 6, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool reports this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-6.litmus @@
\caption{Causality Test Case 6}
\label{lst:Causality Test Case 6}
\end{listing}

A compiler that can prove \co{y} is always either
zero or one can also prove that there is no semantic dependency
from \co{P1()}'s load on line~14 to its stores on lines~16 and~19.
Such a compiler could act as if lines~15--20 had been replaced
by a single store to \co{x}, showing that the dependency isn't
semantic.

Less-omniscient compilers will preserve the dependency, and then
hardware dependency ordering will prevent the OOTA cycle from
being realized.

\subsection{Causality Test Case 7}
\label{app:Causality Test Case 7}

Listing~\ref{lst:Causality Test Case 7}
shows causality test case 7, for which the \co{r1 == r2 == r3 == 1}
result is to be allowed.
The \co{herd7} tool finds this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-7.litmus @@
\caption{Causality Test Case 7}
\label{lst:Causality Test Case 7}
\end{listing}

This test case has everything
to do with reordering and nothing to do with OOTA cycles.
The \co{r1 == r2 == r3 == 1} outcome happens when lines~17, 10, 11,
15, 16 and~9 execute in that order.
Note that this execution order respects the semantic dependencies in
lines~10--11 and~15--16, and thus is not ruled out by hardware
dependency ordering.
It could have been ruled out if some of the accesses were acquire or
release, but here they are all relaxed.

\subsection{Causality Test Case 8}
\label{app:Causality Test Case 8}

Listing~\ref{lst:Causality Test Case 8}
shows causality test case 8, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, but it does if
\co{r2} is initialized to one before being set to
\co{1 + r1 * r1 - r1}.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-8.litmus @@
\caption{Causality Test Case 8}
\label{lst:Causality Test Case 8}
\end{listing}

Omniscient compilers might note that the computation on line~9
will have the value one when the load from \co{x} returns either
zero or one, which are the only values \co{x} and \co{y} can take.
Such a compiler could then realize that line~9 always sets \co{r2} to one,
independent of the value loaded in line~8.
Then there would be no semantic dependency between \co{P0()}'s load
and store, and thus no OOTA cycle relative to this compiler.

A less-capable compiler would leave the dependency in place,
and the resulting hardware dependency ordering would prevent
the OOTA cycle from being realized.

\subsection{Causality Test Case 9}
\label{app:Causality Test Case 9}

Listing~\ref{lst:Causality Test Case 9}
shows causality test case 9, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-9.litmus @@
\caption{Causality Test Case 9}
\label{lst:Causality Test Case 9}
\end{listing}

This is the same as causality test case~8 but with \co{P2()} added;
the store to \co{x} in \co{P2} can interfere with the formation of the
OOTA cycle.

Unlike in causality test case~8, the computation on line~9 must be
carried out because now the load on line~8 might return two.
Consequently \co{P0} and~\co{P1} both have semantic dependencies,
and hardware dependency ordering will prevent the OOTA cycle from
being realized.

\subsection{Causality Test Case 9a}
\label{app:Causality Test Case 9a}

Listing~\ref{lst:Causality Test Case 9a}
shows causality test case 9a, for which the \co{r1 == r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-9a.litmus @@
\caption{Causality Test Case 9a}
\label{lst:Causality Test Case 9a}
\end{listing}

This is the same as causality test case~9 except that here \co{P2()}
sets \co{x} to zero instead of two.
As as result the computation on line~9 is once again independent of
the value loaded on line~8, which makes the analysis nearly the same
as that of causality test case~8.

\subsection{Causality Test Case 10}
\label{app:Causality Test Case 10}

Listing~\ref{lst:Causality Test Case 10}
shows causality test case 10, for which the \co{r1 == r2 == 1} and \co{r3 == 0}
result is to be forbidden.
The \co{herd7} tool reports this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-10.litmus @@
\caption{Causality Test Case 10}
\label{lst:Causality Test Case 10}
\end{listing}

\co{P0()} and \co{P1()} have straightforward
semantic dependencies between their respective loads and stores, and
so hardware dependency ordering will prevent an OOTA cycle from being
realized.
\co{P2()} has no semantic dependency, and \co{P3()}'s store does not
occur in this execution (i.e., the one described by the \co{exists} clause)
because \co{r3} is zero.
Thus neither of them contributes to an OOTA cycle.

\subsection{Causality Test Case 11}
\label{app:Causality Test Case 11}

Listing~\ref{lst:Causality Test Case 11}
shows causality test case 11, for which the \co{r1 == r2 == r3 == r4 == 1}
result is to be allowed.
The \co{herd7} tool reports this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-11.litmus @@
\caption{Causality Test Case 11}
\label{lst:Causality Test Case 11}
\end{listing}

Like causality test case~7, this test case has nothing to do with OOTA cycles.
The result may be obtained by executing lines~20, 12, 13, 18, 19, 10, 11,
and~17 in that order.

\subsection{Causality Test Case 12}
\label{app:Causality Test Case 12}

Causality Test Case 12 in
Listing~\ref{lst:Causality Test Case 12}
uses arrays, which are not yet supported by the \co{herd7} tool.

\begin{listing}[tbp]
@@ DisplayLitmus litmus/oota-causality-12.litmus @@
\caption{Causality Test Case 12}
\label{lst:Causality Test Case 12}
\end{listing}

An omniscient compiler could note that only \co{P0()} accesses
array \co{a[]} and act as if the code was simpler, but there would
still be semantic dependencies between \co{P0()}'s and \co{P1()}s
relaxed atomic loads and stores involving \co{x} and \co{y}.
The resulting hardware dependency ordering will prevent an OOTA cycle
from being realized, whether the compiler is omniscient or not.

\subsection{Causality Test Case 13}
\label{app:Causality Test Case 13}

Listing~\ref{lst:Causality Test Case 13}
shows causality test case 13, for which the \co{r1 == r2 == 1}
result is to be forbidden.
The \co{herd7} tool reports this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-13.litmus @@
\caption{Causality Test Case 13}
\label{lst:Causality Test Case 13}
\end{listing}

This is the same as causality test case~10 without the confounding
influence of \co{P2()} and~\co{P3()}.

\subsection{Causality Test Case 14}
\label{app:Causality Test Case 14}

Listing~\ref{lst:Causality Test Case 14}
shows causality test case 14, for which the \co{r1 == r3 == 1} and \co{r2 == 0}
result is to be forbidden.
The \co{herd7} tool reports this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-14.litmus @@
\caption{Causality Test Case 14}
\label{lst:Causality Test Case 14}
\end{listing}

In the execution described by the \co{exists} clause,
\co{P0()} has a semantic dependency from line~9 to
line~13 and \co{P1()} has a semantic dependency from line~18 to
line~21.
(There's also a semantic dependency from line~17 to line~21 but it
doesn't enter into this OOTA cycle, which involves only \co{x} and~\co{y}.)
The \co{memory_order_seq_cst} specification on line~17 has no important effect;
hardware dependency ordering will prevent the cycle from being realized.

\subsection{Causality Test Case 15}
\label{app:Causality Test Case 15}

Listing~\ref{lst:Causality Test Case 15}
shows causality test case 15, for which the \co{r0 == r1 == r3 == 1}
and \co{r2 == 0}
result is to be forbidden.
The \co{herd7} tool reports this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-15.litmus @@
\caption{Causality Test Case 15}
\label{lst:Causality Test Case 15}
\end{listing}

The execution described by the \co{exists} clause
has an OOTA cycle involving \co{u} and~\co{v}.
The dependency in \co{P0()} from line~13 to line~19 and the dependency
in \co{P1()} from line~24 to line~27 are both semantic.
The resulting hardware dependency ordering prevents the cycle from
being realized.

Having no load, \co{P2()} cannot participate in an OOTA cycle.

\subsection{Causality Test Case 16}
\label{app:Causality Test Case 16}

Listing~\ref{lst:Causality Test Case 16}
shows causality test case 16, for which the \co{r1 == 2} and \co{r2 == 1}
result is to be allowed.
The \co{herd7} tool does not report this result, which is to be
expected because it would violate the C++ memory model's read-write
coherence rule regarding the modification order of an atomic object
(6.9.2.2p17 \co{[intro.races]}).
Because no abstract execution can obtain \co{r1 == 2} and \co{r2 == 1}, no
valid realization will either.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-16.litmus @@
\caption{Causality Test Case 16}
\label{lst:Causality Test Case 16}
\end{listing}

\subsection{Causality Test Case 17}
\label{app:Causality Test Case 17}

Listing~\ref{lst:Causality Test Case 17}
shows causality test case 17, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result, although it does
report a different OOTA cycle involving lines 11, 12, 16, and~17
(the \co{S17} value on line~3 of the \co{herd7} output).

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-17.litmus @@
\caption{Causality Test Case 17}
\label{lst:Causality Test Case 17}
\end{listing}

A compiler that treats atomics as quasi volatile could reason as follows:
Suppose that \co{P0()}'s load from \co{x} on line~8 returns the value 42.
In that case, the second load on line~11 can be merged with the first,
so that it also returns the value 42.
On the other hand, if the load on line~8 returns some other value then
line~10 will store the value 42 to \co{x}, and then the load on line~11
can be omitted in favor of using the value 42 stored by line~10.

Either way, the value of \co{r1} will be 42.
Therefore the loads from \co{x} cannot affect the store on line~17,
so there is no semantic dependency.
This allows the cycle to be realized,
but because the dependency in \co{P0()} is not semantic,
the cycle is not OOTA.

\subsection{Causality Test Case 18}
\label{app:Causality Test Case 18}

Listing~\ref{lst:Causality Test Case 18}
shows causality test case 18, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-18.litmus @@
\caption{Causality Test Case 18}
\label{lst:Causality Test Case 18}
\end{listing}

This is the same as causality test case~17, even to the alternate OOTA
cycle reported by \co{herd7}, except that line~9 tests
\co{r3} for equality to zero rather than non-equality to 42.
Perhaps surprisingly, this makes a difference.

A compiler that treats atomic objects as volatile or uses
single-thread analysis cannot assume that \co{x} will necessarily be
equal to 42 following lines~9 and~10; it might have some other value
such as 1.
Therefore the compiler cannot conclude that \co{r1} will always be 42
and must leave the semantic dependency between lines~11 and~12 intact.
Of course, the resulting OOTA cycle will not be realized, because of
hardware dependency ordering.

\subsection{Causality Test Case 19}
\label{app:Causality Test Case 19}

Listing~\ref{lst:Causality Test Case 19}
shows causality test case 19, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result but it does report the
OOTA cycle in line~3 of its output.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-19.litmus @@
\caption{Causality Test Case 19}
\label{lst:Causality Test Case 19}
\end{listing}

This is nearly the same as the Simple OOTA cycle and causality test
case~4, with the addition of \co{P2()}.
Just as in those examples, the OOTA cycle cannot be realized by
compilers that treat atomic objects as volatile or that do
single-thread analysis and treat atomic objects as quasi volatile.

\subsection{Causality Test Case 20}
\label{app:Causality Test Case 20}

Listing~\ref{lst:Causality Test Case 20}
shows causality test case 20, for which the \co{r1 == r2 == r3 == 42}
result is to be allowed.
The \co{herd7} tool does not report this result.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-causality-20.litmus @@
\caption{Causality Test Case 20}
\label{lst:Causality Test Case 20}
\end{listing}

This is the same as causality test case~19, even to the OOTA cycle
reported by \co{herd7}, except that line~19 tests
\co{r3} for equality to zero rather than non-equality to 42.
The same analysis applies.

>>>>

All OOTA workers owe a debt to the foundational work in the infamous
``Causality Test Cases''.\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
% %%% Eventually cite P3064.

@@

Later examples will usually combine the litmus test and the \co{herd7}
output into one listing, as shown in
Listing~\ref{lst:Simple Reordering as Litmus Test},
which recasts
Listing~\ref{lst:Simple Reordering}
into litmus-test form.
Listing~\ref{lst:Simple OOTA}
can also be recast as a \co{herd7} litmus test, as shown in
Listing~\ref{lst:Causality Test Case 4}
on page~\pageref{lst:Causality Test Case 4}.
Other OOTA-related litmus tests may be found in
Appendix~\ref{app:Illustrative Litmus Tests}.

@@

This paper focuses primarily on compilers that do only local per-thread
analysis.
Further explorations could consider additional classes of global analysis,
for example, analyses that demonstrate that a given expression will
always evaluate to a particular constant.
JMM Causality Test Case 1 % %%% Cite P3064.
% %%% discussed in
% %%% Appendix~\ref{app:Causality Test Case 1}
provides an example of such a global analysis.

@@

The first step in this direction is
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4},
which expands
Listing~\ref{lst:Causality Test Case 4}
% %%% Cite P3064.
% %%% (discussed in
% %%% Appendix~\ref{app:Causality Test Case 4})
from two threads to three.
The \co{herd7} tool reports OOTA values, and
all threads have straightforward semantic dependencies from their
loads to their stores.

@@
